---
# DDEA Annual Day.
title: "Using Training to Move Us Towards More Openness and Reproducibility"
author: "Luke W. Johnston"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: "xaringan-themer.css"
    nature:
      slideNumberFormat: ""
      ratio: "16:9"
      countIncrementalSlides: false
---

layout: true

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
library(emoji)

fa2 <- function(icon, fill = "#da9100") {
  fontawesome::fa(name = icon, fill = fill)
}
```

```{css}
.footer-right {
    background-color: #FFFFFF;
    position: absolute;
    bottom: 10px;
    right: 8px;
    height: 60px;
    width: 30%;
    font-size: 11pt;
}
```

```{r xaringan-themer, include=FALSE}
library(xaringanthemer)
style_mono_accent(
  base_color = "#990033",
  base_font_size = "21px",
  text_font_size = "1.2rem",
  text_font_google = google_font("Quicksand"),
  header_font_google = google_font("Josefin Sans")
)
```

<!-- TODO: convert this to functions? -->

.footer-right[
Slides: [slides.lwjohnst.com/dda/2022-12-01](https://slides.lwjohnst.com/dda/2022-12-01/)
Licensed under CC-BY `r fa2("creative-commons-by", "#990033")`
]

---

class: middle, center

```{r progress-bar}
xaringanExtra::use_progress_bar(color = "#990033", location = "top")
```

```{r logo-sdca}
xaringanExtra::use_logo(
  image_url = "../../common/sdca-logo.png",
  link_url = "https://www.stenoaarhus.dk/",
  width = "60"
)
```

# `r emoji("warning")` Disclaimer `r emoji("warning")`

## Contains advertisement for the courses `r emoji("stuck_out_tongue_winking_eye")`

???

Words ~240

Title slide:

When my wonderful colleagues first mentioned they were going to nominate me for
the award, I was quite nervous because probably like many people, I'm not super
comfortable getting this type of attention. But I was also excited because it
meant I could potentially talk more about the course.

I just want to say, just how awesome it is for there to be an award like this.
I've always talked about how impactful education and networking is, but how
little it often gets recognized as a legitimate research activity.

Because teaching researchers is a very different challenge than teaching
undergrads and in other more formal university courses. You really need to
ground the learning material into something tangibly related to what we as
researchers face and need on a frequent basis. And that requires not just
knowing the science of teaching and learning, but also actually having 
experience actually doing the research. And that takes up a lot of time to
do both. So even if I didn't get this award, I really appreciate that DDA has
created it.

Alright, before I start end up talking about how awesome I think DDA is, I
should move on.

Disclaimer slide:

Like a conflict of interest slide, I felt I should have a disclaimer for this talk,
because really, I thought this would be a great chance to talk about and advertise
the course more! So, this is a warning!

---

# Me, R, and teaching in Canada

--

```{r}
knitr::include_url("https://uoftcoders.github.io/studyGroup/", height = "425px")
```

???

I want to briefly talk a bit about the history of how I got so much into
teaching. During my PhD, I was learning how to do data analysis mostly by
myself, like many many others. I got some help with the other PhD students in
the group, it wasn't very formal. I had an old laptop that wasn't running
Windows Vista very well, so I installed Linux to keep using it for work. Unlike
nowadays where Linux is a very similar experience to Windows, at the time you
had to use some programming to use Linux. So as I was learning data analysis,
I also knew there were ways to use the computer to reduce manual work. But 
everyone who taught me was doing things the very manual way. And I was so
confused, why were we not being taught in courses how to do data analysis more
effectively. As I continued by Masters and PhD, that confusion though, quickly
turned into frustration that we weren't taught this.

That's when I realized, if you want something done, do it yourself. So, with
a few other PhD students from across the university, we started a "coding club"
where peers taught each other computing skills. This club is still running, and 
is a great source of pride for me.

---

# Me, R, and teaching in Denmark

--

```{r}
knitr::include_url("https://au-cru.github.io/site/", height = "425px")
```

???

Words: ~70

Then I moved to Denmark and I wanted to do something similar. I started setting
up this group, called the Aarhus University CRU, but I quickly realized that
Denmark is different from Canada, surprise! I couldn't do the same thing as I
did in Canada. I was talking to my previous supervisor Daniel Witte about teaching,
and he mentioned talking to Tore about running a course through the DDA.

---

# Me, R, and teaching in Denmark

--

.center[
## Enter DDA `r emoji("muscle")` `r emoji("muscle")`
]

???

Words: ~100

So, during EASD, I met with Tore and told him a bit about the idea of the course...
and he didn't look convinced, maybe even a bit confused, why people would take
the course. But, he isn't the type of person to just turn down someone offering
to run a course. We planned for it the next year... and well, it was quite popular
and that's when I think he realized... ok, this is something people want. So after
that first course, he asked "any plans for another, more advanced one"... And 
I said, "hell yes"!

---

```{r, out.width="90%", dev="svg"}
#| fig-width: 8
#| fig-height: 4
#| echo: false
library(tidyverse)
library(lubridate)

intro_courses <- tibble::tibble(
  level = "introduction",
  dates = c("2019-03-04", "2020-06-22", "2021-06-15", "2022-03-01"),
  attendees = c(23, 23, 25, 30),
)

intermediate_courses <- tibble::tibble(
  level = "intermediate",
  dates = c("2020-09-08", "2021-06-28", "2021-10-11", "2022-06-21"),
  attendees = c(20, 23, 19, 30),
  location = rep("DDA", times = 4)
)

bind_rows(
  intro_courses,
  intermediate_courses
) %>%
  mutate(
    attendees_labels = glue::glue("n={attendees}"),
    attendees_y = attendees + 5,
    dates = as_date(dates),
    attendees = if_else(level == "intermediate", -attendees, attendees),
    attendees_y = if_else(level == "intermediate", -attendees_y, attendees_y)
  ) %>%
  ggplot() +
  geom_hline(aes(yintercept = 0)) +
  geom_pointrange(
    aes(x = dates, y = attendees, ymin = 0, ymax = attendees, colour = level),
    size = 0.3
  ) +
  scale_colour_viridis_d(begin = 0.2, end = 0.8, option = "inferno") +
  geom_text(aes(x = dates, y = attendees_y, label = attendees_labels)) +
  coord_cartesian(ylim = c(-45, 45), xlim = as_date(c("2019-01-01", "2022-12-01"))) +
  annotate("text", x = as_date("2020-01-01"), y = 5, label = "Introduction") +
  annotate("text", x = as_date("2020-01-01"), y = -5, label = "Intermediate") +
  theme_minimal() +
  guides(colour = "none") +
  labs(title = "Courses and attendees over the years") +
  theme(
    panel.grid = element_blank(),
    panel.grid.major.x = element_line("grey90", linetype = "dotted"),
    axis.title = element_blank(),
    axis.text.y = element_blank()
  )
```

---

# What makes these courses different?

## Guiding principles:

--

.pull-left[
- Inclusive and open

- Multiple learning activities

- Continually assess and improve
]

--

.pull-right[
- Less is more

- Grounded in real-world practice

- Design for learner *and* instructor
]

--

**It's also lots of fun!**

???

Features of the course

- Openness: Material is online, freely accessible, always.
  - Openly licensed
- Less is more
  - Focused
- Assessing and improving
  - Both material and teaching (from peers)
- Design for both learner and instructor
  - Easier onboarding
- Anchored in real-world practice
  - Focus on how a researcher might *actually* work

---

# What makes these courses different?

## Highly relevant to the modern research environment

- Connect to rise of data science

- Higher demands for openness and verifiability (reproducibility)

- More complex data and analysis problems

---

## Introduction: r-cubed.rostools.org

```{r}
knitr::include_url("https://r-cubed.rostools.org/", height = "500px")
```

---

## Intermediate: r-cubed-intermediate.rostools.org

```{r}
knitr::include_url("https://r-cubed-intermediate.rostools.org/", height = "500px")
```

---

## Guides: guides.rostools.org

```{r}
knitr::include_url("https://guides.rostools.org/", height = "500px")
```

---

## Next step: r-cubed-advanced.rostools.org

```{r}
knitr::include_url("https://r-cubed-advanced.rostools.org/", height = "500px")
```

---

class: middle

## Please use the material! And give me feedback to improve it
