---
# DDEA Annual Day.
title: "Using Training to Shift Towards Openness and Reproducibility"
author: "Luke W. Johnston"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: "xaringan-themer.css"
    nature:
      slideNumberFormat: ""
      ratio: "16:9"
      countIncrementalSlides: false
---

layout: true

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
library(emoji)

fa2 <- function(icon, fill = "#da9100") {
    fontawesome::fa(name = icon, fill = fill)
}
```

```{css}
.footer-right {
    background-color: #FFFFFF;
    position: absolute;
    bottom: 10px;
    right: 8px;
    height: 60px;
    width: 30%;
    font-size: 11pt;
}
```

```{r xaringan-themer, include=FALSE}
library(xaringanthemer)
style_mono_accent(
    base_color = "#990033",
    base_font_size = "21px",
    text_font_size = "1.2rem", 
    text_font_google = google_font("Nanum Gothic"),
    header_font_google = google_font("Merriweather Sans", "300")
)
```

<!-- TODO: convert this to functions? -->

.footer-right[
Slides: [slides.lwjohnst.com/misc/2022-12-01](https://slides.lwjohnst.com/misc/2022-12-01/)
Licensed under CC-BY `r fa2("creative-commons-by", "#990033")`
]

---

```{r progress-bar}
xaringanExtra::use_progress_bar(color = "#990033", location = "top")
```

```{r logo-sdca}
xaringanExtra::use_logo(
  image_url = "../../common/sdca-logo.png",
  link_url = "https://www.stenoaarhus.dk/",
  width = "60"
)
```

class: middle

# `r emo::ji("warning")` Disclaimer `r emo::ji("warning")`

## Contains advertisement for the courses `r emo::ji("stuck_out_tongue_winking_eye")`

???

For first slide, maybe talk a bit about why this is so important to me.

- Teaching researchers is a different challenge compared to undergrads or more formal university courses.
    - Need to really ground the learning into something tangibly related to what we researchers face and need.

Just a small disclaimer...

---

class: middle

# Me, R, and teaching:

##

???

Why wasn't I taught that I don't need to 

---

- History
    - Why started doing R teaching
    - What I did at the University of Toronto
    - Meeting Tore at EASD

---

```{r}
knitr::include_url("https://r-cubed.rostools.org/")
```


---

- Courses ran with DDA
    - Links
        - First did intro, Tore asked about other possible ones, I made an intermediate... then Advanced

```{r}
library(tidyverse)
library(lubridate)
library(ggrepel)

intro_courses <- tibble::tibble(
    level = "introduction",
    dates = c("2019-03-04", "2019-05-20", "2020-06-22", "2021-06-15", "2022-03-01"),
    attendees = c(23, 29, 23, 25, 30),
    location = c("DDA", "DIfE+DZD", "DDA", "DDA", "DDA")
)

intermediate_courses <- tibble::tibble(
    level = "intermediate",
    dates = c("2020-09-08", "2021-06-28", "2021-10-11", "2022-06-21"),
    attendees = c(20, 23, 19, 30),
    location = rep("DDA", times = 4)
)

bind_rows(
    intro_courses,
    intermediate_courses
) %>% 
    mutate(
        dot_labels = glue::glue("{location}\nn={attendees}"),
        dates = as_date(dates),
        location = if_else(location == "DDA", "", location),
        attendees = if_else(level == "intermediate", -attendees, attendees)
    ) %>% 
    ggplot(aes(x = dates, y = attendees, ymin = 0, ymax = attendees, label = dot_labels)) +
    geom_hline(aes(yintercept = 0)) +
    geom_pointrange() +
    geom_text_repel(min.segment.length = Inf) +
    coord_cartesian(ylim = c(-45, 45), xlim = as_date(c("2019-01-01", "2022-12-01"))) +
    annotate("text", x = as_date("2020-01-01"), y = 0, label = "Introduction") +
    theme_minimal() +
    theme(panel.grid = element_blank(),
          panel.grid.major.x = element_line("grey90", size = 1),
          axis.title.y = element_blank(),
          axis.text.y = element_blank())
```


---

# What makes these courses different?

---

- Features of the course
    - Principles?
        - Inclusion
        - Openness
        - Continued involvement
        - Less is more
            - Focused
        - Assessing and improving
            - Both material and teaching (from peers)
        - Design for both learner and instructor
        - It's fun!
    - Anchored in real-world practice
        - Focus on how a researcher might *actually* work


- Or! Use the material yourself to teach your group or others!

1. Collect (and use) anonymous and critical feedback often! Including from instructors
2. Multiple activities to learning *in class* (reading, doing, listening,
discussing, teaching, group, and solo)
3. Strong and focused narrative tied to learning outcome: Less is more
1. Design material for participants *as well as* potential/new instructors 
1. Have material be online, easily found
    - Follow it (almost) exactly, deviate only if necessary
1. Encourage participants to instruct future workshops
    - Provide onboarding documentation and support



