---
title: "Shifting Science and Research Culture Towards Openness and Reproducibility Through Education and Training"
author: "Luke W. Johnston"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: "xaringan-themer.css"
    nature:
      slideNumberFormat: ""
      ratio: "16:9"
      countIncrementalSlides: false
---

layout: true

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)

fa2 <- function(icon) {
    fontawesome::fa(name = icon, fill = "#2471a3")
}

image_link <- function(image_path, url, ...) {
    htmltools::a(
        href = url,
        htmltools::img(src = image_path, ...)
    )
}
```

```{css}
.footer-left {
    background-color: #FFFFFF;
    position: absolute;
    bottom: 8px;
    left: 20px;
    height: 60px;
    width: 30%;
    font-size: 14pt;
}

.footer-right {
    background-color: #FFFFFF;
    position: absolute;
    bottom: 8px;
    right: 5px;
    height: 50px;
    width: 30%;
    font-size: 14pt;
}
```

```{r xaringan-themer, include=FALSE}
library(xaringanthemer)
style_mono_accent(
    base_color = "#2471a3",
    base_font_size = "24px",
    header_font_google = google_font("Merriweather Sans", "300")
)
```

.footer-left[
```{r footer-logos}
image_link("../../common/dda-logo.png", "https://www.danishdiabetesacademy.dk", width = "95px")
image_link("../../common/sdca-logo.png", "https://www.stenoaarhus.dk/", width = "60px")
```
]

.footer-right[
[slides.lwjohnst.com/misc/2021-11-18](https://slides.lwjohnst.com/misc/2021-11-18/)
]

---

```{r progress-bar}
xaringanExtra::use_progress_bar(color = "#A35624", location = "top")
```

```{r logo-r3}
xaringanExtra::use_logo(
  image_url = "https://gitlab.com/uploads/-/system/project/avatar/15345313/apple-touch-icon.png",
  link_url = "https://r-cubed.rostools.org/",
  width = "55",
)
```

# Outline

<!-- 10 min -->

- Key points
- Introduction to course and attendees
- Key principles used while creating/updating course
    - Lessons learned
- Experiences of effectiveness and impact (within Denmark)
- Summary

???

Hi everyone, thanks for coming to listen to my talk. I'm Luke and I work as a
postdoc at the Steno Diabetes Center in Aarhus Denmark. If you want to check out
these slides, there's a link to them in the footer of each slide.

So, since this is an audience of other trainers, I'm going to focus on the
topics of practicals of teaching and on curriculum development.

(Go over outline).

While talking about key principles and lessons learned, it will tie back
to some of the key messages.

---

# *Actionable* key `r emo::ji("key")` points

???

Before I begin, I want to say a couple of caveats about these key messages.
Here, I'm assuming that the learners are working/full-time researchers
(including PhD students), not necessarily in an undergraduate context and
related to learning data analysis or more practical type skills.

--

1. Have material be online, easily found
    - Follow it (almost) exactly, deviate only if necessary
1. Encourage participants to instruct future workshops
    - Provide onboarding docs and support
1. Emphasize code of conduct and safety of the learning space
    - *Embody* it
1. Use and weave in reading, listening, doing, and discussing activities *in class*

???

(Go over the messages).

---

class: middle

# Reproducible Research in R (R3 or r-cubed) course/workshop for PhD students and postdocs *doing biomedical research*

.footnote[
- Website: [r-cubed.rostools.org](https://r-cubed.rostools.org)
- JOSE paper: [10.21105/jose.00122](https://jose.theoj.org/papers/10.21105/jose.00122)
- Also intermediate course: [r-cubed-intermediate.rostools.org](https://r-cubed-intermediate.rostools.org/)
]

???

Now getting into the course itself. In Denmark, a course is anything from a couple 
days to a couple weeks, while in North America this might be referred to as a workshop.
The course is teaching reproducible research in R to PhD students and postdocs who
do biomedical research, largely diabetes research. This course is 3 full days, 
composing of 5 code along sessions where the instructor types and the learners
follow along, a few lectures, and a final group project. For more info on the
course, check out the website as well as our paper describing the course and how
to use it. We also have an intermediate course that has a lot of overlap with
what I'm talking about today.

---

class: middle

# Course arose from need for skills in data analysis `r emo::ji("woman_technologist_medium_skin_tone")`

## ... and for more awareness on reproducibility and open science `r emo::ji("exploding_head")`

???

I started teaching these topics during my Masters and PhD because I saw a huge need
for these skills and just general awareness of some practicalities of doing science.
This course specifically I started during my postdoc because one, there was a need 
for more computational skills in my field and two, because the awareness around
reproducibility and open science was very lacking.

---

class: middle

# Key `r emo::ji("key")` principles when developing material and learned from feedback

???

When developing this course along with updating each time we run it, there
were some key principles that we used to form the foundation for the entire content
of the course. And I'll be covering some of them for most of the talk.

---

## Use evidence based learning and teaching best practices `r emo::ji("woman_teacher_light_skin_tone")`

???

The first one is, of course, building the curriculum based on evidence around learning
and on teaching best practices.

--

1. Collect (and use) anonymous and critical feedback often!

???

Like any research, in order to build and then improve something you need to actively
get feedback, often. Doesn't matter how you do it, but get it. I make a Google Forms
survey that asks generally, what worked and what could be improved, at the end of
each day. Then I use that to determine what needs fixing and what to keep. Must of the
later points link to this item.

--

2. Multiple activities to learning *in class* (reading, doing, listening,
discussing, teaching, group, and solo)

???

Learning requires multiple different angles repeating basically the same information
in a different way until it gets remembered and saved into long-term memory.
In this way, during the course we get learners to do multiple tasks like reading
activities, hands-on exercises, listening, typing while we do the code along,
teaching which I'll get back to, and doing a group project. Most of these
activities essentially repeat the same information but in slightly different
ways.

In relation to reading, one thing I found was talking about a concept always had
some variance in saying it each time I ran the course or when another instructor
did the specific session. Plus, we have many people who have English as a second
language, so the amount of information you could convey orally was much less
than by reading. It also slows us down, which is really important to have enforced
within the material because, I do it, its very easy to go too fast through the material.
After they read a section, as the instructor we would briefly
go over the topic again, to reinforce the concept. After adding the reading
activities and getting the feedback, there was consistent and strong positive comments
about that, so we kept it and added more.

--

3. Less is more: Brains can only handle so much new information

--

4. Strong and focused narrative tied to learning outcome

???

Brains can only hold so much information in them at any one time, so the key
thing while building material is keep things lean and focused on what the learning
outcome and objectives. Both point 3 and 4 relate to this. The narrative throughout
the whole curriculum should be laser focused on the learning outcome. Don't deviate
from the material. Each time we've ran the course, we've stripped out more and more.
Plus, everytime I deviate too much from the material because I have some expert knowledge
that I want to share from excitedness, I see it in the feedback that it confused
them or that if I talk about it it should be in the material.

--

5. Limit lectures and slides (largely for conceptual/ranting topics)


???

Lastly, lectures are some of the poorest form of memory retention, so we severly
limit them to only very abstract topics or things to rant about, like how un-open
science is.

---

class: middle

## Openly licensed `r fa2("fab fa-creative-commons-by")` and easily accessible `r fa2("fas fa-wifi")`

--

.footnote[Website: From Hugo Academic (blogdown) to GitBook (bookdown) to Bootstrap theme (bookdown).]

???

- Re-use for both learners and (potential) instructors

E.g. started with using a typical style website, like that offered by the Hugo
Academic theme, but for several reasons especially with ease of use, I moved
over to using bookdown. The feedback we've gotten

Consistent and strong positive feedback on material being online and easily accessible

- Have material be easily accessible and well designed/structured

6. Book format is very useful since a course is chronological and time is linear,
so book follows exactly what each session should go over and cover.

- I think this is really important because I see a lot of for instance university
courses not have their material be publicly available in the idea that
people should pay for the material and I understand but I think that's a shame
because people don't usually take courses for the content itself, they do it
to be taught. Taught by someone.

---

class: middle

## Safe `r fa2("fas fa-hand-holding-heart")` and supportive `r fa2("fas fa-hands-helping")` environment

.footnote[Through a code of conduct, explicit expectations of participants and
instructors, as well as detailed and descriptive syllabus.]

???

This is really the first step, the
first introduction learners have with the material, so its really important to get
it clear about what they should expect to learn.

- BE VERY MINDFUL of your language. Using words like "just", "simply",
and other "dismissive" language can undermine the learners experience
and comfort, thus impacting their learning.

- STRONGLY emphasize code of conduct, importance of keeping environment
feeling safe and supportive

- Empowering learners to take next steps
    - e.g. Contribute to material via GitLab/Hub

---

class: middle

## Use modern and beginner-friendly software, packages, and workflows

.footnote[E.g. use open source, tidyverse R packages, R Markdown. Small exception `r emo::ji("grimacing")`: Git, though used through RStudio's interface.]

???

- E.g. R and tidyverse, R Markdown. Had a long discussion with a colleague teacher
about only focusing on the tidyverse
- That also fit with philosophy of science (so only open source)

---

class: middle

## Design material for participants *as well as* potential/new instructors 

.footnote[As well as potential contributors (i.e. the participants).]

???

Used GitLab, have GitLab CI build book and website so contributors don't have to, used
bookdown since it has a low skill entry for use (compared to blogdown)


- Provide onboarding documentation, meeting, and support to new instructors
    - Make teaching the course a component of learning (for the new-ish instructor)

---

# Experiences of effectiveness and impact (within Denmark `r emo::ji("denmark")`)

.footnote[Feedback data is found on the [GitLab repository](https://gitlab.com/rostools/r-cubed/-/tree/main/feedback).]

???

- We run the course through the danish diabetes academy here in Denmark, which
provides a training and funding space for diabetes researchers. This course as well
as the intermediate course are both the highest rated and most in-demand courses
that they offer. Plus we can see from the feedback that 

- Every time I teach the course, I ask if they have heard of or know about open
science or reproducibility etc. When I first ran the course, I mostly got blank
stares... now easily more than half know about it in some way.

---

# Summary and recap

1. Have material be online, easily found
    a. Follow it (almost) exactly, deviate only if necessary
1. Encourage participants to instruct future workshops
    a. Provide onboarding documentation and support
1. Emphasize code of conduct and the safety of the space to learn
    a. *Embody* it
1. Use and weave in reading, listening, doing, and discussing *in class*

--

### Try out the material and give us feedback on it! `r emo::ji("nerd")`

???

# Appendix

Why not use existing material from e.g. Carpentries?

- No strong narrative thread
- Jumping around with concepts without reusing them again later
- Using older workflows/packages
- Emphasis/focus on writing software
- Not targeted to biomedical researchers
