---
title: "Shifting Science and Research Culture Towards Openness and Reproducibility Through Education and Training"
author: "Luke W. Johnston"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: "xaringan-themer.css"
    nature:
      slideNumberFormat: ""
      ratio: "16:9"
      countIncrementalSlides: false
---

layout: true

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)

fa2 <- function(icon) {
    fontawesome::fa(name = icon, fill = "#2471a3")
}

image_link <- function(image_path, url, ...) {
    htmltools::a(
        href = url,
        htmltools::img(src = image_path, ...)
    )
}
```

```{css}
.footer-left {
    background-color: #FFFFFF;
    position: absolute;
    bottom: 8px;
    left: 20px;
    height: 60px;
    width: 30%;
    font-size: 14pt;
}

.footer-right {
    background-color: #FFFFFF;
    position: absolute;
    bottom: 8px;
    right: 5px;
    height: 50px;
    width: 30%;
    font-size: 14pt;
}
```

```{r xaringan-themer, include=FALSE}
library(xaringanthemer)
style_mono_accent(
    base_color = "#2471a3",
    base_font_size = "24px",
    header_font_google = google_font("Merriweather Sans", "300")
)
```

.footer-left[
```{r footer-logos}
image_link("../../common/dda-logo.png", "https://www.danishdiabetesacademy.dk", width = "95px")
image_link("../../common/sdca-logo.png", "https://www.stenoaarhus.dk/", width = "60px")
```
]

.footer-right[
[slides.lwjohnst.com/misc/2021-11-18](https://slides.lwjohnst.com/misc/2021-11-18/)
]

---

```{r progress-bar}
xaringanExtra::use_progress_bar(color = "#A35624", location = "top")
```

```{r logo-r3}
xaringanExtra::use_logo(
  image_url = "https://gitlab.com/uploads/-/system/project/avatar/15345313/apple-touch-icon.png",
  link_url = "https://r-cubed.rostools.org/",
  width = "55",
)
```

# Outline

<!-- 10 min -->

- Key points
- Introduction to course and attendees
- Key principles used while creating/updating course
    - Lessons learned
- Experiences of effectiveness and impact (within Denmark)
- Summary

???

Hi everyone, thanks for coming to listen to my talk. I'm Luke and I work as a
postdoc at the Steno Diabetes Center in Aarhus Denmark. If you want to check out
these slides, there's a link to them in the footer of each slide.

So, since this is an audience of other trainers, I'm going to focus on the
topics of practicals of teaching and on curriculum development.

(Go over outline).

While talking about key principles and lessons learned, it will tie back
to some of the key messages.

---

# *Actionable* key `r emo::ji("key")` points

???

Before I begin, I want to say a couple of caveats about these key messages.
Here, I'm assuming that the learners are working/full-time researchers
(including PhD students), not necessarily in an undergraduate context and
related to learning data analysis or more practical type skills.

--

1. Have material be online, easily found
    - Follow it (almost) exactly, deviate only if necessary
1. Encourage participants to instruct future workshops
    - Provide onboarding documentation and support
1. Emphasize code of conduct and safety of the learning space
    - *Embody* it
1. Use and weave in reading, listening, doing, and discussing activities *in class*

???

(Go over the messages).

---

class: middle

# Reproducible Research in R (R3 or r-cubed) course/workshop for PhD students and postdocs *doing biomedical research*

.footnote[
- Website: [r-cubed.rostools.org](https://r-cubed.rostools.org)
- JOSE paper: [10.21105/jose.00122](https://jose.theoj.org/papers/10.21105/jose.00122)
- Also intermediate course: [r-cubed-intermediate.rostools.org](https://r-cubed-intermediate.rostools.org/)
]

???

Now getting into the course itself. In Denmark, a course is anything from a couple 
days to a couple weeks, while in North America this might be called a workshop.
The course is teaching reproducible research in R to PhD students and postdocs who
do biomedical research, largely diabetes research. 

This course is 3 full days, composing of 5 code along sessions where the
instructor types and the learners follow along, a few lectures, and a final
group project. For more info on the course, check out the links below.

---

class: middle

# Course arose from need for skills in data analysis `r emo::ji("woman_technologist_medium_skin_tone")`

## ... and for more awareness on reproducibility and open science `r emo::ji("exploding_head")`

???

While I've been teaching these general topics since my Masters, this course
specifically I started during my postdoc because one, there was a need for more
computational skills in my field and two, because the awareness around
reproducibility and open science was very lacking.

---

class: middle

# (Some) Key `r emo::ji("key")` principles when developing the material and learned from the feedback

???

When developing this course along with updating each time we run it, there
were some key principles that we used to form the foundation for the entire content
of the course. And I'll be covering some of them for most of the talk.

---

## Use evidence based learning and teaching best practices `r emo::ji("woman_teacher_light_skin_tone")`

???

The first one is, of course, building the curriculum based on evidence around learning
and on teaching best practices.

--

1. Collect (and use) anonymous and critical feedback often! Including from instructors

???

Like any research, in order to build and then improve something you need to actively
get feedback, often. I made a Google Forms survey that asks generally, what
worked and what could be improved, at the end of each day. Then I use that to
fix, remove, or keep things.

--

2. Multiple activities to learning *in class* (reading, doing, listening,
discussing, teaching, group, and solo)

???

Learning requires multiple different angles repeating basically the same information
in a different way until it gets remembered and saved into long-term memory.
In this way, during the course we get learners to do multiple tasks like reading
activities, hands-on exercises, listening, and so on. 

In relation to reading, one thing I found was talking about a concept always had
some variance in saying it each time I ran the course or when another instructor
did the specific session. Plus, we have many people who have English as a second
language, so the amount of information you could convey orally was much less
than by reading. 

It also slows us down, which is really important to have enforced within the
material because, I do it, its very easy to go too fast through the material.
After they read a section, as the instructor we would briefly go over the topic
again, to reinforce the concept. After adding the reading activities and getting
the feedback, there was consistent and strong positive comments about that, so
we kept it and added more.

--

3. Strong and focused narrative tied to learning outcome: Less is more

???

The narrative throughout the whole curriculum should be laser focused on the
learning outcome.

Brains can only hold so much information in them at any one time, so the key
thing while building material is keep things lean and focused on what the
learning outcome and objectives. Each time we've ran the course, we've stripped
out more and more.

Don't deviate from the material. Plus, everytime I deviate too much from the
material because I have some expert knowledge that I want to share from
excitedness, I see it in the feedback that it confused them or that if I talk
about it it should be in the material.

--

4. Limit lectures and slides (largely used for conceptual/ranting topics)


???

Lastly, lectures are some of the poorest form of memory retention, so we severely
limit them to only very abstract topics or things to rant about, like how un-open
science is. 

---

class: middle

## Openly licensed `r fa2("fab fa-creative-commons-by")` and easily accessible `r fa2("fas fa-wifi")`

???

The next key principle is to ensure the material is maximally used and re-used.
This means applying an open license to it and having it easily accessible online.

I think this is a bit of a misunderstood area. I see a lot of for instance university
courses that don't have their material publicly available, thinking that they 
need to keep it only for people paying or attending the class. While I understand 
that idea, I think that's a shame and maybe misguided because people don't
usually take courses for the content itself, generally, they take them to be
taught. They come for teachers and the experience, not the content alone.

And we see it in the feedback, there is always very strong and consistent
positive feedback talking about how great it is that the material is online and
easily accessible.

---

class: middle

## Safe `r fa2("fas fa-hand-holding-heart")` and supportive `r fa2("fas fa-hands-helping")` environment

.footnote[Through a code of conduct, explicit expectations of participants and
instructors, as well as detailed and descriptive syllabus.]

???

This next principle is something I think is so important... and that is establish
and maintaining an environment that is safe and supportive for learning. Have the
explicit code of conduct on expected behaviour of everyone involved at the forefront and 
have instructors embody this attitude. Learning is so dependent on the learners
motivation and interest and engagement with the content... and if they don't feel
safe, they won't learn as much as they could.

And it's so easy to unconsciously use dismissive language like saying
"simply" or hand-waving a topic away when you're teaching. But it can lead to
undermining learners confidence and trust in themselves, particularly for
underrepresented or minority groups within science. So constantly self-reflecting
is really vital to this topic.

For example, I have a question in the pre-course survey asking
if they have read the code of conduct, which is also a pre-course task.
Then during the first introduction session, I mention that I've read that
everyone say yes to that question and I make it clear, this is important, this
isn't something on the side.

---

class: middle

## Design material for participants *as well as* potential/new instructors 

.footnote[As well as potential contributors (i.e. the participants).]

???

Lastly, the material isn't only for the participants, it's also for future instructors
and contributors. Since it is written with the idea that learners will read it
during the course or afterwards, this also gives new instructors something to 
use to prepare for teaching their session or course itself if they want.
Interspersed throughout the text are messages to the instructors on top of
an instructor guide giving advice and tips on how to prepare, what to do,
and how to teach. We hope that, structured as it is, anyone can take the course
and try to teach it themselves to others.

---

class: middle

# Experiences of effectiveness and impact (within Denmark `r emo::ji("denmark")`)

.footnote[Feedback data is found on the [GitLab repository](https://gitlab.com/rostools/r-cubed/-/tree/main/feedback).]

???

Alright, I've gone over the key principles for the course, how effective and impactful
has it been? We've ran the course through the Danish Diabetes Academy here in
Denmark, which is an organization that provides funding, networking events, and
educational events and courses to diabetes researchers. They collect evaluations
for the course and both this course as well as the intermediate one are both
the highest rated and most in-demand courses that they offer. Plus based on
the feedback we collect, we can see how much the participants really enjoy the
course and often recommend it their colleagues. If you want to read some of the 
feedback we get, they are found in our GitLab repository, the link is there in
the footnote of this slide.

As for impact, every time I teach the course, I ask if they have heard of or
know about open science or reproducibility etc. When I first ran the course, I
mostly got blank stares... now easily more than half know about it in some way.
It probably isn't only due to the course, in general this area along with data
science are growing so quick that people would have eventually heard about these
topics.

---

# Summary and recap

1. Have material be online, easily found
    - Follow it (almost) exactly, deviate only if necessary
1. Encourage participants to instruct future workshops
    - Provide onboarding documentation and support
1. Emphasize code of conduct and the safety of the space to learn
    - *Embody* it
1. Use and weave in reading, listening, doing, and discussing *in class*

???

Anyway, I could talk about this for a long time. But to end, let's go back to
the key messages. I talked about (go over messages).

--

### Try out the material and give us feedback on it! `r emo::ji("nerd")`

???

And to end, please, if you try out the material, lets us know! We'd love more
feedback on it! Thanks for listening!

# Appendix

Why not use existing material from e.g. Carpentries?

- No strong narrative thread
- Jumping around with concepts without reusing them again later
- Using older workflows/packages
- Emphasis/focus on writing software
- Not targeted to biomedical researchers
