---
title: "Parquet, DuckDB, and fastreg for blazingly fast data analysis in the Danish Registers"
date: "2026-06-23"
date-format: long
author: Luke W. Johnston
location: "Epi group meeting at Steno Diabetes Center Aarhus, Denmark (on the 23rd) and for Department of Public Health, Aarhus University (on the 24th)"
---

# Outline of talk {.center}

-   Rationale and aim
-   Parquet: A modern format for efficiently storing and loading data
-   DuckDB: Fast and powerful data processing and analysis
-   fastreg: Convert SAS to Parquet and quickly load registers

# Rationale for this talk {.center}

Recent developments spark a push for better tooling

## Big picture: Academia is woefully behind in using and knowing modern computing tools {.center}

## This impacts our ability to do research effectively and productively {.center}

## Working with massive data has its own challenges and needs {.center}

## DST recently requires limiting RAM and resource usage {.center}

## DST charges for storage space {.center}

## Aim of this talk {.center}

1.  Spread knowledge of better practices in structuring and working with data.

2.  Showcase fastreg to improve how to work more effectively on DST.

# Parquet file format {.center}

<https://parquet.apache.org/>

## Parquet is a column-based data storage format

Most data formats are row-based, like CSV. Newer formats tend to be
column-based.

::: columns
::: column
### Row-based

``` text
name,gender,age
Tim,Man,30
Sam,Woman,30
Jenny,Woman,25
```
:::

::: {.column .fragment}
### Column-based

``` text
name,Tim,Sam,Jenny
gender,Man,Woman,Woman
age,30,30,25
```
:::
:::

## Column-based storage has many advantages

::: columns
::: column
### Compression

``` text
name,Tim,Sam,Jenny
gender,Man,Woman,Woman
age,30,30,25
```

...becomes...

``` text
name,Tim,Jenny,Sam
gender,Man,Woman{2}
age,30{2},25
```
:::

::: {.column .fragment}
### Loading

-   Computers read by lines
-   Per line = same data type
-   Only read needed columns

``` text
gender,Man,Woman,Woman
age,30,30,25
```
:::
:::

## Parquet is 50-75% smaller than other formats {.center}

| File type            | Size (MB)    |
|----------------------|--------------|
| SAS (`.sas7bdat`)    | 1.45 Gb      |
| CSV (`.csv`)         | \~90% of SAS |
| Stata (`.dta`)       | 745 Mb       |
| Parquet (`.parquet`) | 398 Mb       |

: File size between CSV, Parquet, Stata, and SAS for `bef` register for
2017. Note: SAS files on DST are bigger than normal.

## Can partition data by a value (e.g. year) and load all at once {.center}

Called "Hive partitioning":

``` text
bef/
├── year=2020/
│   └── part-0.parquet
├── year=2021/
│   └── part-0.parquet
└── year=2022/
    └── part-0.parquet
```

## SAS and Python can load Parquet but not Stata {.center}

([StataNow](https://www.stata.com/statanow/) can, but isn't on DST...)

Should use R anyway :wink:

::: notes
Not sure if StataNow can load in a Hive Partitioned Parquet dataset.
:::

# DuckDB {.center}

::: aside
<https://duckdb.org/>
:::

## DuckDB is a fairly recent SQL engine designed for analytical queries {.center}

## SQL, or Structured Query Language, is a language for managing and querying databases {.center}

## Is impressively fast {.center}

-   Faster than almost all other tools.

    -   Relatively complex queries (joins, group by, aggregates) on 55
        Gb takes \< 7.5 min [^1].

    -   Generally, simpler queries take \< 10 seconds for massive
        datasets.

-   Easily connects with Parquet datasets.

[^1]: From benchmark results
    [page](https://duckdblabs.github.io/db-benchmark/). Some tools couldn't even handle the queries.

## Can be easily used in R {.center}

``` r
library(arrow)
bef <- open_dataset("path/to/bef") |>
  to_duckdb()

# Or can use duckplyr
library(duckplyr)
bef <- read_parquet_duckdb("path/to/bef/*/*.parquet")
```
## Integrates with tidyverse and its syntax {.center}

``` r
library(tidyverse)
bef |>
  filter(year = 2020) |>
  select(pnr, koen)
```

## SAS recently added DuckDB support, Stata has community packages for it {.center}

(Not sure if either are on DST though.)

Should use R anyway :wink:

# fastreg

<https://dp-next.github.io/fastreg/>

::: notes
For those familiar with the dstDataPrep, we took that code out of DST and made
it into a more structured, better designed package that we published to CRAN.
:::

## Key features {.center}

- Converts SAS into Hive partitioned Parquet datasets

- Automatically handle data updates from DST and Externe registers

- Can do it in parallel for faster conversion

- Easily load all years of a register at once and connect to DuckDB, with minimal use of memory

## Don't use StatTransfer on DST, use fastreg {.center}

StatTransfer doesn't create a Hive partition! :astonished:

## Comes with targets pipeline template for easy setup {.center}

Recommended to do it this way, create with `fastreg::use_template()`.

## A simple demo (without targets) {.center}

Need to set location of SAS files and eventual Parquet output location:

```{r}
#| label: setup
#| include: false
# pak::pak("dp-next/fastreg")
library(tidyverse)
library(furrr)

show_tree <- function(dir) {
  withr::with_dir(fs::path_dir(dir), {
    fs::dir_tree(fs::path_file(dir))
  })
}
```

```{r}
#| label: set-options
#| echo: true
options(
  # With a fake project ID and the temporary directory.
  fastreg.project_rawdata_dir = fs::path_temp("E/rawdata/701020/"),
  fastreg.project_workdata_dir = fs::path_temp(
    "E/workdata/701020/parquet-registers/"
  )
)
```

## Converting these SAS files ... {.center .smaller .scrollable}

```{r}
#| label: simulate-to-sas
#| echo: false
#| warning: false
#| message: false
rawdata_dir <- getOption("fastreg.project_rawdata_dir")
registers_tbl <- fastreg::simulate_registers_with_paths(
  registers = c("bef", "lmdb"),
  years = as.character(2020:2025),
  n = 1000000,
  output_dir = rawdata_dir
)

plan(multisession)
sas_paths <- registers_tbl |>
  furrr::future_pwalk(fastreg::write_to_sas) |>
  dplyr::pull(output_path)

# list_sas_files(rawdata_dir) |>
#   fs::file_delete()

rawdata_dir |>
  # Get to the `E` directory to list both SAS and Parquet files
  fs::path_dir() |>
  fs::path_dir() |>
  show_tree()
```

## ... Using this fastreg code ... {.center .smaller .scrollable}

(note: targets template has better parallel processing)

```{r}
#| label: convert-to-parquet
#| message: false
#| echo: true
library(tidyverse)
library(furrr)
# Where SAS files are.
rawdata_dir <- getOption("fastreg.project_rawdata_dir")
# Set where Parquet should be saved.
workdata_dir <- getOption("fastreg.project_workdata_dir")
plan(multisession)
conversion_log <- rawdata_dir |>
  fastreg::list_sas_files() |>
  future_map(\(path) fastreg::convert(path, output_dir = workdata_dir)) |>
  list_rbind()
```

## ... Into these Parquet datasets {.center .smaller .scrollable}

```{r}
#| label: list-parquet-workdata
#| echo: false
workdata_dir |>
  # Get to the `workdata` directory to list both SAS and Parquet files
  fs::path_dir() |>
  fs::path_dir() |>
  show_tree()
```

## ... And you can quickly list all registers as datasets {.center}

*Requires the `options()` to be set.*

``` r
fastreg::list_parquet_datasets()
```

```{r}
#| label: list-parquet-datasets
#| echo: false
fastreg::list_parquet_datasets() |>
  stringr::str_remove("^/tmp/[[:alnum:]]+/")
```

## Parquet is much smaller than SAS, saving on storage costs {.center}

```{r}
#| label: parquet-vs-sas-size
#| echo: false
sas_size <- sum(fs::file_info(sas_paths)$size) |>
  as.character()
parquet_size <- fastreg::list_parquet_files() |>
  fs::file_info() |>
  pull(size) |>
  sum() |>
  as.character()
```

- All SAS files: `r sas_size` bytes
- All Parquet files: `r parquet_size` bytes

*Note: SAS files are larger than normal in DST, so this difference will be bigger.*

## Reading all years of a register at once {.center .smaller .scrollable}

(available for anyone working in that DST project.)

```{r}
#| label: set-workdata-dir
#| eval: false
#| echo: true
library(fastreg)
options(
  fastreg.project_workdata_dir = "E:/workdata/701010/parquet/"
)
bef <- read_register("bef")
lmdb <- read_register("lmdb")
```

```{r}
#| label: show-bef
#| echo: false
fastreg::read_register("bef") |>
  select(-source_file) |>
  print(n = 4)
```

## Tests comparing simple analysis fastreg vs direct from SAS {.center .smaller}

How long does it take to do a simple count of gender in each 5 years in `bef` (1 million rows)?

1. Read in all the data for all the years
2. Group by year
3. Summarise number of people by gender
4. Arrange by year, koen, number

## Using fastreg (with Parquet and DuckDB) is blazing fast! {.center}

```{r}
#| label: convert-to-rds
#| include: false
# Save to Rds so we can compare between fastreg, Rds, and SAS.
rds_dir <- workdata_dir |>
  fs::path_dir() |>
  fs::path_temp("rds")
fs::dir_create(rds_dir)
save_to_rds <- function(path, output_dir) {
  rds_file <- path |>
    fs::path_file() |>
    fs::path_ext_set("rds")

  haven::read_sas(path) |>
    saveRDS(fs::path(output_dir, rds_file))
}
sas_paths |>
  purrr::walk(\(path) save_to_rds(path, rds_dir))
fs::dir_ls(rds_dir, recurse = TRUE)
```

```{r}
#| label: benchmark-sas-vs-fastreg
#| echo: false
#| warning: false
read_sas_with_years <- function(path) {
  year <- fs::path_file(path) |>
    fs::path_ext_remove() |>
    stringr::str_extract("\\d{4}")
  path |>
    haven::read_sas() |>
    dplyr::mutate(year = as.integer(year))
}

test_sas <- function(paths) {
  paths |>
    stringr::str_subset("bef") |>
    purrr::map(read_sas_with_years) |>
    purrr::list_rbind() |>
    dplyr::summarise(n = n(), .by = c(year, koen)) |>
    dplyr::arrange(year, koen, n) |>
    dplyr::ungroup() |>
    tibble::as_tibble()
}

test_fastreg <- function() {
  fastreg::read_register("bef") |>
    dplyr::summarise(n = n(), .by = c(year, koen)) |>
    dplyr::arrange(year, koen, n) |>
    dplyr::collect() |>
    tibble::as_tibble()
}

read_rds_with_years <- function(path) {
  year <- fs::path_file(path) |>
    fs::path_ext_remove() |>
    stringr::str_extract("\\d{4}")
  readRDS(path) |>
    dplyr::mutate(year = as.integer(year))
}

test_rds <- function(dir) {
  rds_paths <- dir |>
    fs::dir_ls(recurse = TRUE, regexp = "bef")
  rds_paths |>
    purrr::map(read_rds_with_years) |>
    purrr::list_rbind() |>
    dplyr::summarise(n = n(), .by = c(year, koen)) |>
    dplyr::arrange(year, koen, n) |>
    tibble::as_tibble()
}

bench::mark(
  "From SAS" = test_sas(sas_paths),
  "From RDS" = test_rds(rds_dir),
  fastreg = test_fastreg(),
  iterations = 10
) |>
  dplyr::select(
    "Approach" = expression,
    "Median time" = median,
    "Memory used" = mem_alloc
  ) |>
  knitr::kable()
```

([Code available here](https://github.com/lwjohnst86/slides/blob/main/s/2026-06-23/index.qmd))

## Limitations of DuckDB {.center}

- Can't do advanced statistical analysis

- A few tidyverse functions don't work

**Solution: Process with DuckDB, collect into R, then do stats**

## Check out the fastreg website for more details on how to use it {.center}

<https://dp-next.github.io/fastreg/>
